\\Large Rising temperatures alter how and where mammals move

### 2025-10-03 writing/manuscript.Rmd
addressed more of the comments from KH

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness and community dynamics. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but few disentangled them from seasonal behaviour cycles. Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21^st^ century. We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species (boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves) throughout western Canada between 1998 and 2023. We show that temperature significantly affected how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century, although species did not respond uniformly to temperature. Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less. Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in seasonal phenology. We also predict that rising temperatures will reduce median habitat selection strength for most of the six species, but some are expected to show increased selection strength for higher altitudes. As mammals increasingly alter their movement rates and select against portions of their current ranges, changes in individuals' movement behaviour will impact encounter rates, including predator-prey dynamics and human-wildlife interactions. Conservation efforts should therefore account for future changes in movement behaviour as well as the consequences such changes may have on communities. Anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. How and where animals move thus also determines encounter rates with resources [@mezzini_how_2025], humans [@johnson_human_2018; @weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can impact animals' movement rates [@dyer_travel_2023], phenology [@carbeck_adaptation_2022; @johnson_human_2018; @mclellan_effect_2015; @pigeon_drivers_2016], and life history [@newediuk_climate_2024], and even cause physiological damage and death [@hetem_responses_2014; @powers_hovering_2017; @ratnayake_forecasting_2019; @schmidt_interplay_2020; @schwerdt_ecological_2024]. Animals may search for a short-term buffer from extreme temperatures by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016]. However, climate change during the last century [@ipcc_climate_2023] has caused animals to alter how and where they move as they respond to hotter temperatures and increasingly severe and frequent extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011].

@@

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves; Table \@ref(tab:data-table)). Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme heat events [@zhang_increased_2023]. Decision makers must understand which areas will best sustain and protect biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].

```{r bc-map, fig.cap="GPS telemetry data for the six species in this study (dots) over a digital elevation model for the study area.", out.width='0.75\\linewidth'}
knitr::include_graphics('../figures/tels-map-dem.png')
```

@@

\noindent Of the GPS fixes with finite speed estimates, 2.6% had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:hgam-summary), Fig. B2). Sampling interval affected estimates of probability as well of as speed when moving (Fig. B8). Consequently, we present all results while predicting specifically for one-hour sampling intervals. At $0^\circ$C, species differed in estimated mean probabilities of moving ($\hat P(M = 1)$; range: 0.05 -- 0.31), mean speed when moving ($\hat {\mathbb E}(S|M=1)$; range: 0.42 -- 2.67 km/h), and mean overall speed (i.e., $\hat P(M) \times \hat {\mathbb E}(S|M=1)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:hgam-summary)). Grizzly bears had the lowest movement frequency ($\hat P(M) \approx 0.05$), while wolves and cougars moved most often ($\hat P(M) \ge 0.22$). Mountain goats and southern mountain caribou moved the slowest ($\hat{\mathbb E}(S|M = 1) \approx 0.43$ km/h), while wolves had the highest mean speed when moving ($\hat{\mathbb E}(S|M = 1) \approx 2.67$ km/h). Consequently, at $0^\circ$C, wolves traveled an average of $0.22 \times 2.67~\text{km/h} \approx 0.6$ km/h; 2.5 to 16.7 times further than other species.

@@

intercepts <- tibble(
  Species = SPECIES,
  p = get_int(readRDS('models/binomial-gam.rds')),
  s = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
  d = p * s) %>%
  rename('P(M=1|T=0^\\circ C)' = p,
         'E(S|M=1,T=0^\\circ C)' = s,
         'E(D|T=0^\\circ C)' = d) %>%
  mutate(Species = COMMON_NAMES) %>%
  arrange(Species) %>%
  rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
  rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
  rename_with(\(.x) gsub('P', '\\\\hat{P}', .x)) %>%
  rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x))

@@

# species-level proportions
d %>%
  rename(Species = species) %>%
  group_by(Species) %>%
  summarize('$n$' = n(),
            'T<-20\\textdegree{}C (%)' = get_perc(temp_c < -20),
            'T>+20\\textdegree{}C (%)' = get_perc(temp_c > +20)) %>%
  mutate(Species = COMMON_NAMES[map_int(Species,
                                        \(.s) which(SPECIES == .s))]) %>%
  full_join(intercepts, by = 'Species') %>%
  arrange(Species) %>% # arrange alphabetically by common name
  mutate(Species = if_else(Species == 'Caribou (southern mountain)',
                           'Caribou (s. mountain)', Species)) %>%
  # add overall proportions
  bind_rows(
    d %>%
      summarize(Species = 'Total',
                '$n$' = n(),
                'T<-20\\textdegree{}C (%)' = get_perc(temp_c < -20),
                'T>+20\\textdegree{}C (%)' = get_perc(temp_c > +20))) %>%
  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = "Summary statistics for each species' GPS fixes with finite speed estimates, namely: the number fixes after data cleaning ($n$), the percentage of fixes with temperature ($T$) below $-20^\\circ$C and above $20^\\circ$C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S|M=1)$; km/h), and the mean hourly distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S|M=1) = \\hat{\\mathbb E}(D)$; km/h), for a sampling interval of 1 hour and a temperature $T$ of $0^\\circ$C.")
```

@@

\noindent Species' changes in movement rates to temperature varied in both direction and magnitude (Figs. \@ref(fig:dist), B4-B6), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs. B4-B6). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models had good in-sample prediction (Fig. B7) and explained reasonably high proportions of the deviance [79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses]. All species altered their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). The response was most visible in cougars. In late spring (June 1^st^), they moved from evening to early morning if hourly temperatures were below $20^\circ$C, but if temperatures were above 20$^\circ$C they moved mostly between 3:00 and 6:00 AM. Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" was relative to the time of year. All species' estimated probability of moving and speed when moving decreased with sampling intervals above 1 hour, except for cougars' speed, although the estimated trends were highly uncertain (Fig. B8). Uncertainties around the estimated effects were generally higher at extreme temperatures due to lower data availability (Figs. B4A, B5A, and B6A).

@@

```{r hrsfs, fig.cap="\\textbf{Mammals' habitat selection depends on ambient temperature.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature. The grey contours indicate the extent of each species' observed data. RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
knitr::include_graphics('../figures/hrsf-surface-plots.png')
```

@@

Median RSS was projected to decrease over time within the each species' observed range, but, again, changes were stronger under worse SSPs (Fig. \@ref(fig:t-rss-cc)). Decreases were most pronounced in areas with the lowest RSS and most severe for elk, mountain goats, pumas, and southern mountain caribou. Changes for boreal wolves and caribou were relatively negligible. Elk, cougars, and grizzly bears were predicted to increase their selection strength for top-RSS areas, and elk, mountain goats, cougars, and southern mountain caribou were predicted to further decrease their selection for areas with low RSS. Unsurprisingly, the change in RSS between 2025 and 2100 also varied spatially for all species (Fig. \@ref(fig:s-rss-cc)). Overall, RSS decreased throughout most of each species' current range, although elk, cougars, and bears were predicted to increase their selection for higher-altitude habitats. Still, none of the species were projected to increase RSS throughout their habitat (\@ref(fig:density-hrsfs)).

@@

\noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that it affects mammals' seasonal and daily movement behaviour in complex and nonlinear ways. However, predicting mammals' responses to climate change remains a challenge, as habitats are warming rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015]. We leveraged the flexibility and interpretability of HGAMs to estimate mammals' nonlinear responses to temperature without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature with data from multiple individuals. By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into windows (e.g., day/night and seasons). We were thus able to present results using relatively simple and interpretable surface plots (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed nuanced, continuous-time estimates of change. This framework was essential for estimating how mammals behave at extreme temperatures. In fact, despite the recent increase in frequency of extreme heat events in western Canada [@zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by GAMs [@wood_generalized_2017]. Additionally, hierarchical models allowed us to produce statistically sound estimates that accounted for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully. While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].

Species varied greatly in how they adapted their movement behaviour to temperature. For boreal caribou, hotter temperatures increased movement rates in fall, winter, and spring but decreased them in summer, mirroring previous studies [@stien_congruent_2012; @leclerc_determinants_2021; @lessard_local_2025]. The increase in movement rates with temperature during the cold months is likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but warmer temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025]. The reduced movement rates during warmer summers are likely mainly due to more frequent resting and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023]. Therefore, although we predict that boreal caribou's yearly average movement rates will remain approximately unchanged throughout the 21^st^ century, the projections do not show the changes in seasonal movement phenology. This is an important caveat, since the effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction. Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer to encroach on boreal caribou habitat and increase the density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023], whose movement behaviour also depends on ambient temperature. Boreal wolves responded to temperature similarly to boreal caribou, and habitat selection strength was not projected to change noticeably by 2100, but wolves were predicted to move less in future decades. In contrast, cougars showed markedly different seasonal responses to temperature, with warmer temperatures always resulting in less travel, which resulted in strong predicted decreases in movement rates over the 21^st^ century. However, spatial heterogeneity in habitat, climate, and cougar physiology and behaviour may cause cougars in boreal regions to move differently from those in our study. Still, rising temperatures clearly affect more than each individual species' movement behaviour; they have complex and interconnected effects on entire communities and their trophic structures.

@cunningham_opportunity_2021 discuss the importance of adopting a community-level perspective when assessing the effects of climate change on animals' movement behaviour and phenology. For example, as temperatures warm and both boreal caribou and wolves increase their movement rates in fall, winter, and spring, they also increase their encounter rates [@martinez-garcia_how_2020], which will likely increase predation risk when caribou are pregnant or with young calves. Additionally, while both species saw reduced movement rates during hot summer days, boreal wolves' use of anthropogenic linear features (e.g., roads, seismic lines) may allow them to reduce the thermal costs of movement while increasing the chances of encountering heat-stressed prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022]. Furthermore, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may still risk increasing predation pressure from bears [@leblond_caribou_2016] and the aforementioned encroaching predators [@demars_incorporating_2023; @labadie_global_2023].

Considerations about changes in trophic interactions leads to another important caveat about the results we present. The estimated changes in movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature. For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be in part due to increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023]. However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025]. Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species' movement rates, but it does not explicitly account for energetic costs. Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs. Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside each species' study area. For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different elevations.

>>>>

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness and community dynamics. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but few disentangled them from seasonal behaviour cycles. Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21^st^ century. We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species (boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves) throughout western Canada between 1998 and 2023. We show that temperature significantly affected how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century. Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less. Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in seasonal phenology. We also predict that rising temperatures will reduce median habitat selection strength for four of the species, although cougars and elk are expected to show increased selection strength for higher altitudes. As mammals increasingly alter their movement rates and select against portions of their current ranges, changes in individuals' movement behaviour will impact encounter rates, including predator-prey dynamics and human-wildlife interactions. Conservation efforts should therefore account for future changes in movement behaviour as well as the consequences such changes may have on communities. Anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. How and where animals move thus also determines encounter rates with resources [@mezzini_how_2025], humans [@johnson_human_2018; @weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can impact animals' movement rates [@dyer_travel_2023], phenology [@carbeck_adaptation_2022; @johnson_human_2018; @mclellan_effect_2015; @pigeon_drivers_2016], and life history [@newediuk_climate_2024], and even cause physiological damage and death [@hetem_responses_2014; @powers_hovering_2017; @ratnayake_forecasting_2019; @schmidt_interplay_2020; @schwerdt_ecological_2024]. Animals may search for a short-term buffer from extreme temperatures by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016]. However, climate change during the last century [@ipcc_climate_2023] has caused animals to alter how and where they move as they respond to hotter temperatures and increasingly severe and frequent extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011; @thompson_2021_2022].

@@

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves; Table \@ref(tab:data-table)). Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [plants: @post_climate_2008; @tysor_countergradient_2025; ice: @basu_phenological_2024], and more frequent and intense extreme heat events [@zhang_increased_2023]. Decision makers must understand which areas will best sustain and protect biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].

```{r bc-map, fig.cap="GPS telemetry data for the six species in this study (dots) over a digital elevation model for the study area. The inset in the bottom left shows the telemetry data within the north american continent. Both maps use the BC Albers Equal Area Conic projection (EPSG:3005).", out.width='\\linewidth'}
knitr::include_graphics('../figures/tels-map-dem.png')
```

@@

\noindent Of the GPS fixes with finite speed estimates, 2.6% had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:hgam-summary), Fig. B2). Sampling interval affected estimates of probability as well of as speed when moving (Fig. B8). All species' estimated probability of moving and speed when moving decreased with sampling intervals above 1 hour, except for cougars' speed, although the estimated trends were highly uncertain (Fig. B8). Consequently, we present all results while predicting specifically for one-hour sampling intervals. At $0^\circ$C, species differed in estimated mean probabilities of moving ($\hat P(M = 1)$; range: 0.05 -- 0.31), mean speed when moving ($\hat {\mathbb E}(S|M=1)$; range: 0.42 -- 2.67 km/h), and mean overall speed (i.e., $\hat P(M) \times \hat {\mathbb E}(S|M=1)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:hgam-summary)). Grizzly bears had the lowest movement frequency ($\hat P(M) \approx 0.05$), while wolves and cougars moved most often ($\hat P(M) \ge 0.22$). Mountain goats and southern mountain caribou moved the slowest ($\hat{\mathbb E}(S|M = 1) \approx 0.43$ km/h), while wolves had the highest mean speed when moving ($\hat{\mathbb E}(S|M = 1) \approx 2.67$ km/h). Consequently, at $0^\circ$C, wolves traveled an average of $0.22 \times 2.67~\text{km/h} \approx 0.6$ km/h; 2.5 to 16.7 times further than other species.

@@

intercepts <- tibble(
  Species = SPECIES,
  p = get_int(readRDS('models/binomial-gam.rds')),
  s = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
  d = p * s) %>%
  rename('P(M=1|T)' = p,
         'E(S|M=1,T)' = s,
         'E(D|T)' = d) %>%
  mutate(Species = COMMON_NAMES) %>%
  arrange(Species) %>%
  rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
  rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
  rename_with(\(.x) gsub('P', '\\\\hat{P}', .x)) %>%
  rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x))

@@

# species-level proportions
d %>%
  rename(Species = species) %>%
  group_by(Species) %>%
  summarize('$n$' = n(),
            'T<-20\\textdegree{}C (%)' = get_perc(temp_c < -20),
            'T>+20\\textdegree{}C (%)' = get_perc(temp_c > +20)) %>%
  mutate(Species = COMMON_NAMES[map_int(Species,
                                        \(.s) which(SPECIES == .s))]) %>%
  full_join(intercepts, by = 'Species') %>%
  arrange(Species) %>% # arrange alphabetically by common name
  mutate(Species = if_else(Species == 'Caribou (southern mountain)',
                           'Caribou (s. mountain)', Species)) %>%
  # add overall proportions
  bind_rows(
    d %>%
      summarize(Species = 'Total',
                '$n$' = n(),
                'T<-20\\textdegree{}C (%)' = get_perc(temp_c < -20),
                'T>+20\\textdegree{}C (%)' = get_perc(temp_c > +20))) %>%
  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = "Summary statistics for each species' GPS fixes with finite speed estimates, namely: the number fixes after data cleaning ($n$), the percentage of fixes with temperature ($T$) below $-20^\\circ$C and above $20^\\circ$C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S|M=1)$; km/h), and the mean hourly distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S|M=1) = \\hat{\\mathbb E}(D)$; km/h), for a sampling interval of 1 hour and a temperature of $T = 0^\\circ$C.")
```

@@

\noindent Species' changes in movement rates to temperature varied in both direction and magnitude (Figs. \@ref(fig:dist), B4-B6), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs. B4-B6). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models had good in-sample prediction (Fig. B7) and explained reasonably high proportions of the deviance [79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses]. All species altered their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). The response was most visible in cougars. In late spring (June 1^st^), they moved from evening to early morning if hourly temperatures were below $20^\circ$C, but if temperatures were above 20$^\circ$C they moved mostly between 3:00 and 6:00 AM. Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" was relative to the time of year. Overall, uncertainty around the estimated effects was generally higher at extreme temperatures due to lower data availability (Figs. B4A, B5A, and B6A).

@@

```{r hrsfs, fig.cap="\\textbf{Mammals' habitat selection depends on ambient temperature.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature. The grey contours indicate the extent of each species' observed data. RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability."}
knitr::include_graphics('../figures/hrsf-surface-plots.png')
```

@@

Median RSS was projected to decrease over time within each species' observed range, but, again, changes were stronger under worse SSPs (Fig. \@ref(fig:t-rss-cc)). Decreases were most pronounced in areas with the lowest RSS and most severe for elk, mountain goats, cougars, and southern mountain caribou. Changes for boreal wolves and caribou were negligible. Elk, cougars, and grizzly bears were predicted to increase their selection strength for top-RSS areas, and elk, mountain goats, cougars, and southern mountain caribou were predicted to further decrease their selection for areas with low RSS. Unsurprisingly, the predicted change in RSS between 2025 and 2100 also varied spatially for all species (Fig. \@ref(fig:s-rss-cc)). Overall, RSS decreased throughout most of each species' current range, although elk, cougars, and bears were predicted to increase their selection for higher-altitude habitats. Still, none of the species were projected to increase RSS throughout their habitat (\@ref(fig:density-hrsfs)).

@@

\noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that it affects mammals' seasonal and daily movement behaviour in complex and nonlinear ways. However, predicting mammals' responses to climate change remains a challenge, as habitats are warming rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015]. We leveraged the flexibility and interpretability of HGAMs to estimate mammals' nonlinear responses to temperature without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature with data from multiple individuals. By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into windows (e.g., day/night and seasons). We were thus able to present results using relatively simple and interpretable surface plots (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed nuanced, continuous-time estimates of change. This framework was essential for estimating how mammals behave at extreme temperatures. In fact, despite the recent increase in frequency of extreme heat events in western Canada, including a severe heat wave in 2021 [@zhang_increased_2023; @thompson_2021_2022; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by HGAMs [@wood_generalized_2017]. Additionally, the hierarchical approach allowed us to leverage data from multiple individuals in a single model while accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully. While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are as large as those presented here [@wood_generalized_2017; @wood_generalized_2017-1].

The effects of warming temperatures on movement rates varied across species, seasons, and space due to differences in species' behavior and physiology as well as their habitats' phenology. For boreal caribou, hotter temperatures decreased movement rates summer but increased them otherwise, mirroring previous studies [@stien_congruent_2012; @leclerc_determinants_2021; @lessard_local_2025]. The increase in movement rates with temperature during the cold months is likely partly due to shallower snow depth, which results in greater mobility [@pedersen_quantifying_2021], but warmer temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025]. The reduced movement rates during warmer summers are likely mainly due to more frequent resting and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023]. Therefore, although we predict that boreal caribou's yearly average movement rates will remain approximately unchanged throughout the 21^st^ century, the projections do not show the changes in seasonal movement phenology. This is an important caveat, since the effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction. Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer to encroach on boreal caribou habitat and increase the density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023], whose movement behaviour also depends on ambient temperature.

Boreal wolves responded to temperature similarly to boreal caribou, and habitat selection strength was not projected to change noticeably by 2100, but wolves were predicted to move less in future decades. In contrast, cougars showed markedly different seasonal responses to temperature, with warmer temperatures always resulting in less travel, which resulted in strong predicted decreases in movement rates over the 21^st^ century. However, spatial heterogeneity in habitat, climate, and cougar behaviour may cause boreal individuals to respond to temperature differently from those in our study. Still, rising temperatures clearly affect more than each individual species' movement behaviour; they have complex and interconnected effects on entire communities and their trophic structures.

@cunningham_opportunity_2021 discuss the importance of adopting a community-level perspective when assessing the effects of climate change on animals' movement behaviour and phenology. For example, as temperatures warm and both boreal caribou and wolves increase their movement rates in fall, winter, and spring, they also increase their encounter rates [@martinez-garcia_how_2020], which will likely increase predation risk when caribou are pregnant or with young calves. Additionally, while both species saw reduced movement rates during hot summer days, boreal wolves' use of anthropogenic linear features (e.g., roads, seismic lines) may allow them to reduce the total thermal costs of movement by moving for shorter periods while increasing the chances of encountering heat-stressed prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022]. Furthermore, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may still risk increasing predation pressure from bears, cougars, and coyote [@leblond_caribou_2016; @demars_incorporating_2023; @labadie_global_2023].

Considerations about changes in trophic interactions leads to another important caveat about the results we present. The estimated changes in movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature. For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be in part due to increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023]. However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025]. Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species' movement rates, but it does not explicitly account for energetic costs. Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs. Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside the study areaes used here. For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not account for differences in forage availability at different elevations.
